---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# seekr

<!-- badges: start -->

[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/seekr)](https://CRAN.R-project.org/package=seekr)

<!-- badges: end -->

**seekr** is an R package designed to help you search for specific patterns within text files.

## Installation

You can install the development version of **seekr** from GitHub with:

```r
# install.packages("devtools")
devtools::install_github("smartiing/seekr")
```


## Functions

**seekr** provides two main functions:

- `seek()`: Search for a pattern in files within a specified directory.
- `seek_in()`: Search for a pattern in a given list of files.

Each function returns a tibble with the following columns:

- `file`: Index of the file in the list.
- `path`: Path to the file.
- `line`: Line number where the pattern was found.
- `content`: Content of the matching line.
- `match`: The first match found in the line.
- `matches`: All matches found in the line.


## Example

Here's a basic example of how to use **seekr**:

```{r}
# Search for lines containing 'particular words' in csv files within the specified folder
tmp_mtcars = tempfile("01mtcars", fileext = ".csv")
tmp_iris = tempfile("02iris", fileext = ".csv")

write.csv(mtcars, tmp_mtcars)
write.csv(iris, tmp_iris)

found = seekr::seek(
  pattern = "(?i)toyota|honda|setosa", 
  path = tempdir(), 
  filter = "\\.csv$"
)
  
print(found)
  
unlink(c(tmp_mtcars, tmp_iris))
```


## License

This package is licensed under the MIT License.
