---
title: "Design choices"
description: >
  Why seekr represents search results as match vectors, and how this design
  makes search, inspection, filtering, and safe replacement part of the same R workflow.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The central object in `seekr` is the `seekr_match` vector. Technically, it is
implemented as a [`vctrs` record-style vector](https://vctrs.r-lib.org/reference/new_rcrd.html):
a vector object made of several same-length fields, where the internal field
structure supports the object but is not the main user-facing abstraction.

This article describes the main design choices behind `seekr`: how search
results are represented, how they can be inspected and refined, and how
replacement can remain explicit and safe.

## Design requirements

The design started from the workflow `seekr` was meant to support. The goal was
to let users:

- read files once during the search step, and read them again only if
  replacements are applied;
- print matches and planned replacements with surrounding context;
- preview replacements before files are written;
- prepare replacements either during the search step or after matches have been
  inspected;
- update replacements after the search;
- filter matches after the intial search;
- replace only the remaining matches, safely.

More generally, the result needed to expose the information captured for each
match, so users could use it for their own workflows. This includes understanding
what was matched and where, filtering matches more precisely, summarizing search
results, or preparing and updating replacements using the context and other captured fields.

The `seekr_match` vector is the object that came out of these requirements. It
allows workflows like this:

```{r, eval = FALSE}
x <- seekr("old_name", "new_name")

summary(x)
print(x, context = 2)
x <- filter_match(x, path != "legacy.R")
field(x, "replacement") <- toupper(field(x, "replacement"))

replace_files(x)
```

## Why a match vector?

Search-and-replace is often treated as a single operation: find all occurrences
of a pattern and replace them immediately.

`seekr` splits this into separate steps. A user can search first, inspect the
matches, filter out false positives, update replacements, and only then write
changes to disk.

That workflow requires the search result to remain usable as an object after the
search. It must be possible to subset it, combine it, print it, summarize it,
convert it to a tibble, update planned replacements, and eventually pass it to
`replace_files()`.

This is why `seekr` represents search results as a vector of matches.

Each element of the vector represents one match. The vector abstraction makes the
common operations simple: subsetting keeps matches, concatenating combines
matches, and filtering removes matches. The object can behave like a vector while
still storing the metadata needed for inspection and replacement.

## Designs considered

Before settling on the `seekr_match` vector, a few alternative designs were
considered. The two most important were returning a data frame and returning a
file-centric object.

### Why not a data frame?

A data frame was the most obvious alternative. It is familiar to R users, easy to
inspect, and useful for tabular workflows.

`seekr` supports this workflow through `as_tibble()` and `as_match()`, so users
can move to and from a tibble when that representation is convenient. See
[Tabular workflows](tabular-workflows.html) for examples.

But a data frame is not the right core representation for `seekr`.

Most fields in a match should not be edited directly by the user. Fields such as
`path`, `start`, `end`, `match`, `line`, `encoding`, and `hash` describe what was
found. They are part of the recorded search result. In normal use, the only field
users are expected to modify is `replacement`.

In a data frame, all columns feel equally editable. A `seekr_match` vector makes
the intended abstraction clearer: the object is a collection of matches, not a
regular table of analysis variables.

Printing is also not naturally tabular. The useful output for a search result
groups matches by file, shows context, highlights the matched text, and previews
planned replacements.

Finally, a `seekr_match` vector can preserve invariants that would be easier to
break in a plain data frame. It can behave like a vector while still keeping the
fields required for safe replacement together.

### Why not a file-centric object?

Another possible design was a file-centric object: one object per file, with the
matches nested inside each file.

This would mirror the structure of the input files, and it has some attractive
properties. Search happens in files, so grouping by file is natural. Information
shared by all matches in a file, such as the path, encoding, or full text, could
also live once at the file level.

But having a list of file objects as the result of a search would make the 
user-facing workflow harder to understand and to manipulate. Filtering individual 
matches would require working inside a nested structure. Updating replacements 
would mean modifying matches inside file objects. Combining results from several 
searches would raise questions about how file-level objects should be merged.

It would also make common match-level questions more complex. For example,
keeping only matches whose surrounding context contains a specific word should be
a simple filtering operation. In a file-centric object, context would either need
to live inside nested matches, or be recomputed from file-level text. Both options
make the object harder to reason about.

`seekr` treats each match as the unit of work. Files still matter, and matches
still know which file they came from, but the object itself is match-centric. If a
search returns 50 matches across 20 files, the result is a vector of 50 matches.
This makes common operations simpler: subset matches, concatenate matches, filter
by matched text or context, update replacements, and replace the selected
matches.


## Replacing inspected matches safely

The match vector also matters when files are modified.

Many search-and-replace tools show matches first, then run a second search when
replacement is applied. `seekr` does not work that way. `replace_files()` starts
from the current `seekr_match` vector and only considers the matches still
present in that object.

If the user filters out a match, that match is not replaced. If the user changes
the `replacement` field for one match, that match gets its own replacement. The
replacement step follows the object the user inspected and edited.

To make this safe, `seekr` records a hash of the searched text when a
`seekr_match` vector is created. When replacements are applied, the hash of the
current text is computed again: `replace_files()` re-reads the file before
writing, and `replace_text()` checks the text supplied by the user.

If the hash has changed, `seekr` refuses to apply the replacements. This is
strict by design: the goal is not to guess whether recorded positions are still
probably safe, but to ensure that replacements are applied only to the exact text
that was searched and inspected.

The consequence is simple: if the text changed, search again and create a new
`seekr_match` vector. This also means that `seekr` is not designed for sources
that are continuously changing, such as log files receiving new lines every few
seconds. This is an acceptable trade-off for a workflow focused on inspection,
control, and safe replacement.


## Encoding

Encoding is another place where `seekr` chooses an explicit design.

When files are searched, the encoding used to read each file is recorded in the
`seekr_match` vector. When replacements are applied with `replace_files()`,
modified files are written as UTF-8.

By default, `replace_files()` refuses to silently rewrite a non-UTF-8 file as
UTF-8. The user has to make that choice explicitly.

This is a trade-off. `seekr` does not try to be a universal
encoding-preserving editor. Instead, it makes encoding changes visible and
deliberate.

If the user needs full control over reading and writing text, they can use the
text-level workflow described in [Working with text](working-with-text.html).
