---
title: "Getting started with seekr"
description: >
  A detailed walkthrough of the main seekr workflow.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

source_dir <- system.file("extdata", package = "seekr")
example_dir <- file.path(tempdir(), "seekr-example")
backup_dir <- file.path(tempdir(), "seekr-backups")

unlink(example_dir, recursive = TRUE)
unlink(backup_dir, recursive = TRUE)

dir.create(example_dir, recursive = TRUE)
dir.create(backup_dir, recursive = TRUE)

options(seekr.backup_dir = backup_dir)

file.copy(
  from = source_dir,
  to = example_dir,
  recursive = TRUE
)

path <- file.path(example_dir, "extdata")
knitr::opts_knit$set(root.dir = path)
```

## Introduction

`seekr` turns search-and-replace across files into an explicit workflow.

This article gives a detailed walkthrough of that workflow. It starts from the basic steps—listing files, filtering files, and finding matches—and then shows how matches can be inspected, filtered, updated, replaced, and restored.

The examples use a temporary copy of the example files shipped with `seekr`, so they are safe to modify.

```{r, echo = TRUE, eval = FALSE}
library(seekr)
```

```{r, include = FALSE, eval = TRUE}
library(seekr)
options(width = 150)
```

## List files

The first step is to list candidate files.

`list_files()` lists the files that exist and that will be filtered before looking for matches. Its main arguments are `path`, which defines where to look, `all`, which controls whether hidden files and directories are included, and `recurse`, which controls whether and how deeply subdirectories are searched.

By default, `list_files()` searches recursively from the current directory, and
ignores hidden files and directories.

When searching inside a Git repository, you can also set `use_git = TRUE` to
restrict file discovery to files Git considers relevant: tracked files and
untracked files that are not ignored by Git.

```{r list_files}
files <- list_files()
files
```

`list_files()` is intentionally simple. Its role is to discover candidate files, not to decide which files are relevant for a particular search. That decision is handled later by `filter_files()`.

This separation is deliberate. File discovery can stay broad and predictable, while file filtering remains explicit and inspectable.

## Filter files

Once files have been listed, `filter_files()` excludes the files that should not be searched.

### Built-in filters

`filter_files()` takes a vector of file paths, typically returned by `list_files()`, and returns the subset of files that should be searched.

It has three main built-in filters:

- `extension`, to keep files with selected extensions,
- `path_pattern`, to keep files whose path matches a pattern,
- `max_file_size`, to keep files below a size limit.

In addition to these built-in filters, `filter_files()` also uses a list of `exclude_functions`. These are predicate-like functions that decide whether specific files should be excluded. By default, they are used to exclude files that should usually not be searched, such as non-text files or files handled by the default exclusion rules.

For example, we can keep only R files.

```{r filter_files1}
filter_files(files, extension = "R")
```

Here we combine the different types of filters to exclude the files we are not interested in.

```{r filter_files2}
# Add a dummy png file to illustrate the exclusion of non-text files by default
files <- c(files, "server.png")

filtered <- filter_files(
  files, 
  extension = c("r", "log", "yaml", "png"),
  path_pattern = "script|server",
  max_file_size = 1000L
)

filtered
```

### Inspect exclusions

Filtering is inspectable. `filter_files()` records which files were excluded and why.

In this example, files are first excluded by `extension`. Then, among the remaining files, paths that do not match `path_pattern` are excluded. Finally, the two log files are excluded because they are above the size limit set with `max_file_size`.

Each column in the exclusions table represents one filtering or exclusion step. The last column shows that the dummy `"server.png"` file was excluded by the default rules because its extension is not associated with text files.

```{r exclusions1}
exclusions(filtered)
```

This is useful because file filtering can otherwise be hard to audit. Instead of silently excluding files, `seekr` lets you inspect what was excluded at each step.

### Custom exclude functions

In addition to the built-in filters, `filter_files()` uses a list of exclude functions. These functions receive file paths and return `TRUE` for files that should be excluded.

`seekr` provides a default set of exclude functions.

```{r}
names(exclude_functions)
```

You can add your own **named** exclude function by modifying a copy of `exclude_functions`. Custom functions will also appear on their dedicated column in the `exclusions()` data frame.

```{r exclude_functions}
my_exclude_functions <- c(
  exclude_functions,
  exclude_script2 = function(path) grepl("script2[.]R$", path)
)

filtered_custom <- filter_files(
  files,
  extension = "R",
  exclude = my_exclude_functions
)

filtered_custom
```

The order of exclude functions matters. They are evaluated in the order in which they appear in `exclude`, so changing that order can change which exclusion reason is recorded first. It can also matter for performance: computationally expensive functions are usually better placed near the end.

If no exclude functions are used, the exclusions table simply contains the file paths and an `excluded` column set to `FALSE`.

```{r exclusions3}
files |> filter_files(exclude = NULL) |> exclusions()
```

## Find matches

Now that we have selected the files we want to search, we can look for matches with `match_files()`.

The two central arguments are `pattern` and `replacement`. `pattern` defines what to look for. `replacement` is optional: when provided, it prepares replacement values for later, but it does not modify any file. Files are only modified when `replace_files()` is called.

`match_files()` also has a few arguments that control how files are read and how matches are recorded:

- `context` controls how many lines before and after each match are stored in the `seekr_match` vector.
- `encoding` controls how files are decoded. By default, files are read as UTF-8. Set `encoding = NULL` if you want `seekr` to try to detect the encoding with `stringi::stri_enc_detect()`.

### Regular expression patterns

Here we look for function names composed of two words separated by an underscore followed by ` <- function`.

```{r match1}
matches <- match_files(filtered, "([a-z]+)_([a-z]+)(?= <- function)")
matches
```

A plain character string is automatically treated as an ICU regular expression via `stringr::regex()`, as in `stringr`, except that `multiline` is set to `TRUE` by default. Because `multiline = TRUE`, anchors such as `^` and `$` can match line boundaries inside a file.

These two calls are therefore equivalent.

```{r match2, eval = FALSE}
match_files(filtered, pattern)

match_files(
  filtered,
  stringr::regex(
    "([a-z]+)_([a-z]+)(?= <- function)",
    ignore_case = FALSE,
    multiline = TRUE,
    comments = FALSE,
    dotall = FALSE
  )
)
```

For more control, pass a `stringr` pattern object directly. For example, we can make the search case-insensitive.

```{r match3}
match_files(filtered, stringr::regex("FUNCTION", ignore_case = TRUE))
```

For more details about regular expressions in `stringr`, see the [stringr regular expressions documentation](https://stringr.tidyverse.org/articles/regular-expressions.html).

### Literal text patterns

If you want to search for literal text instead of a regular expression, use `stringr::fixed()`.

```{r match4}
match_files(filtered, stringr::fixed("<- function("))
```

This is useful when the text contains characters that would otherwise have a special meaning in a regular expression. Note that `stringr::fixed()` performs bytewise matching, so `stringr::coll()` may be more appropriate for locale-sensitive text.

For more details about the different pattern engines in `stringr`, see the [stringr introduction](https://stringr.tidyverse.org/articles/stringr.html#engines).

## Prepare replacements

Searching and replacing are separate steps in `seekr`.

You can prepare replacements when searching, but files are not modified until you call `replace_files()`. This makes it possible to search first, inspect the result, filter matches, update replacements if needed, and only then write changes to disk.

`seekr` offers five ways to prepare replacements during the matching step:

- `NULL`, the default, to plan no replacement.
- A plain string, used literally as replacement text.
- A string with backreferences of the form `\\1`, `\\2`, and so on, replaced with the corresponding capture groups from `pattern`.
- A function, called once per file with a character vector of all matches found in that file, and expected to return a character vector of the same length, such as `toupper()`.
- A function wrapped with `with_capture_groups_matrix()`, called once per file with a character matrix where the first column is the full match and the remaining columns are the capture groups.

### Literal replacements

The simplest replacement is a string.

```{r replacement1}
match_files(
  filtered,
  pattern = stringr::fixed("safe"),
  replacement = "checked"
)
```

Here, every match of `"safe"` gets the same planned replacement: `"checked"`.

### Capture groups

Replacement strings can also refer to capture groups from a regular expression.

Here, we reverse the two parts of each function name.

```{r replacement2}
match_files(filtered, "(\\w+)_(\\w+)(?= <- function)", "\\2_\\1")
```

### Function replacements

A replacement can also be a function. The function receives a character vector of matched texts and must return one replacement value per match. It is vectorized over the matches found in a file; it is not called separately for each individual match.

```{r replacement3}
match_files(filtered, "([a-z]+)_([a-z]+)(?= <- function)", toupper)
```

### Capture group matrix replacements

For more complex replacements, wrap a function with `with_capture_groups_matrix()`.

The function receives the capture group matrix returned by the matching engine. The first column is the full match, and the following columns are the capture groups.

```{r replacement4}
repl_with_matrix <- function(M) {
  # full_match <- M[, 1L]
  group1 <- M[, 2L]
  group2 <- M[, 3L]
  
  paste0(toupper(group2), "_", tolower(group1))
}

repl_with_matrix <- with_capture_groups_matrix(repl_with_matrix)
match_files(filtered, "([a-z]+)_([a-z]+)(?= <- function)", repl_with_matrix)
```

This is useful when the replacement logic depends on several captured parts of the match.

For workflows where text has already been read, or where you want to control reading and writing yourself, see [Working with text](working-with-text.html).

## Use `seek()` and `seekr()`

The previous sections used the lower-level steps explicitly. For common workflows, `seek()` combines listing, filtering, and matching in one call. `seekr()` is a shortcut around `seek()` with R, R Markdown, and Quarto extensions selected by default.

```{r seek}
files <- list_files()
filtered <- filter_files(files, extension = "R")
x <- match_files(filtered, "(\\w+)_(\\w+)(?= <- function)", "\\2_\\1")

y <- seekr("(\\w+)_(\\w+)(?= <- function)", "\\2_\\1")

identical(x, y)
```

Note that:

- the vector of matches returned by `seekr()` also contains the `exclusions()` attribute,
- when no matches are found, the vector returned by `seek()`/`seekr()` also contains an `empty_stage()` attribute that helps explain where the workflow became empty.

## Inspect the `seekr_match` vector

The result of `match_files()`, `seek()`, and `seekr()` is a `seekr_match` vector.

A `seekr_match` vector behaves like a vector of matches, but each match also stores fields such as the file path, match location, matched text, planned replacement, surrounding context, and encoding.

```{r inspect1}
str(x)
```

You can inspect the available fields with `fields()`.

```{r inspect2}
fields(x)
```

Individual fields can be accessed with `field()`.

```{r inspect3}
field(x, "match")
field(x, "replacement")
field(x, "path")
```

Use `summary()` to get a compact overview of the matches and planned replacements.

```{r summary}
summary(x)
```

Use `print()` to inspect matches with surrounding context and preview replacements before modifying files.

```{r print}
print(x, context = c(2L, 1L))
```

In terminals that support OSC8 hyperlinks, printed file locations can also be clickable.

For more on why `seekr` represents search results this way, see [Design choices](design-choices.html).

## Filter matches

Because a `seekr_match` is a vector, it can be subset like any other R vector.

```{r subsetting}
x[!grepl("safe", field(x, "match"))]
```

However, `filter_match()` is usually more convenient. It evaluates expressions directly on the fields of the `seekr_match` vector.

```{r filter_match}
xf <- x |> filter_match(!grepl("safe", match))
xf
```

The result is still a `seekr_match` vector, so it can be printed, summarized, modified, or passed to `replace_files()`.

```{r print_filter_match}
print(xf, context = c(2L, 1L))
```

## Use tabular workflows

For many workflows, `filter_match()` and `field()` are enough.

For more complex operations, it can be useful to convert a `seekr_match` vector to a tibble, work with it using tabular tools, and then convert it back to a `seekr_match` vector before replacing files.

This is especially useful for grouped summaries, joins, group-aware filtering, or replacement logic that is easier to express with `dplyr`.

When converting back, `as_match()` validates that the tibble still contains the fields required to reconstruct a valid `seekr_match` vector. This makes it possible to use tabular workflows while still returning to the main `seekr` replacement workflow.

For examples, see the `as_match()` documentation and the article [Tabular workflows](tabular-workflows.html).

## Update replacements after inspection

You do not need to decide every replacement when searching.

The `replacement` field can be set or updated after matches have been inspected or filtered.

```{r update_replacement}
first_path <- field(xf, "path")[[1L]]
repl <- field(xf, "replacement")

field(xf, "replacement") <- ifelse(
  field(xf, "path") == first_path,
  toupper(repl),
  tolower(repl)
)

print(xf, context = c(2L, 1L))
```

This makes it possible to search broadly, inspect the result, keep only the matches that matter, and then decide what each selected match should become.

## Replace selected matches

**Important:** `replace_files()` writes modified files in UTF-8. If you need to control how files are written, use the workflow described in [Working with text](working-with-text.html).

When the selected matches and replacements look right, we can call `replace_files()` which starts from the current `seekr_match` vector.

If you found five matches, filtered the vector down to three matches, and updated their replacements, only those three remaining matches are replaced, each with its corresponding replacement.

Before writing, `replace_files()` checks that every selected match has a replacement and that the hash of each affected file still matches the hash recorded when the `seekr_match` vector was created. If a file has changed since the search, replacement stops before writing that file, and the search should be run again on the current file contents.

If you created several `seekr_match` vectors from the same file state, combine them before replacing. After the first replacement, the file hash changes, so a second call using matches from the old file state will fail. For example, use `replace_files(c(x, y))` instead of calling `replace_files(x)` and then `replace_files(y)`.

By default, `backup = TRUE`, so each file that will be modified is backed up before it is written. A `description` can be provided to make the backup easier to identify later. The backup directory can also be changed with `backup_dir` or the `seekr.backup_dir` option.

```{r replace_files}
replaced <- replace_files(xf, description = "Inverse function names")
replaced
```

After replacement, we can search again.

```{r}
seekr("([a-z]+)_([a-z]+)(?= <- function)") |> print(context = c(0L, 3L))
```

## Restore files

After modifying files, backups can be inspected with `list_backups()`, and the most recent backup can be retrieved with `last_backup()`.

```{r}
bck <- last_backup()
bck
```

Use `restore_files()` to restore the previous file contents.

```{r}
restore_files(
  from = bck$backup, 
  to = bck$original,
  description = "restore after reversing function names by mistake"
)
```

By default, restoring files also creates a backup before writing, so both operations remain available in the backup history. This makes it possible to undo a replacement, while still keeping a record of the files that were present just before the restore operation.

```{r}
list_backups()
```

After restoring, the original matches are back.

```{r}
after_restore <- seekr("([a-z]+)_([a-z]+)(?= <- function)")
print(after_restore, context = c(2L, 1L))
```

```{r cleaning, include = FALSE, eval = FALSE}
unlink(example_dir, recursive = TRUE)
unlink(backup_dir, recursive = TRUE)
```
