---
title: "Tabular workflows"
description: >
  Use dplyr and tibble workflows to inspect, filter, and update seekr_match vectors.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

source_dir = system.file("extdata", package = "seekr")
example_dir = file.path(tempdir(), "seekr-example")
backup_dir = file.path(tempdir(), "seekr-backups")

unlink(example_dir, recursive = TRUE)
unlink(backup_dir, recursive = TRUE)

dir.create(example_dir, recursive = TRUE)
dir.create(backup_dir, recursive = TRUE)

options(seekr.backup_dir = backup_dir)

file.copy(
  from = source_dir,
  to = example_dir,
  recursive = TRUE
)

path = file.path(example_dir, "extdata")
knitr::opts_knit$set(root.dir = path)
```

## Overview

`seek()`/`seekr()` returns matches as a `seekr_match` vector. For many workflows, you can inspect and refine this vector directly with `print()`, `summary()`, and `filter_match()`.

Sometimes, however, it is useful to switch to a tabular workflow. This is especially true when you want to summarize matches by file, filter based on group-level information, or prepare replacements with data frame tools.

This article shows how to:

- convert a `seekr_match` vector to a data frame,
- inspect matches with grouped summaries,
- filter matches using group-level logic,
- update replacements with `dplyr`,
- convert back to a `seekr_match` vector with `as_match()`,
- apply replacements with `replace_files()`.

The examples use a temporary copy of the example files shipped with `seekr`, so they are safe to modify.

## Find matches

Suppose we want to inspect log messages.

```{r, echo = TRUE, eval = FALSE}
library(seekr)
library(dplyr)
library(stringr)
```

```{r, include = FALSE, eval = TRUE}
library(seekr)
library(dplyr)
library(stringr)
options(width = 150)
```

```{r}
x <- seek("([A-Z]+) : (.+$)", extension = "log")
x
```

We summarize directly to have an idea of what was found.

```{r}
summary(x)
```

## Convert to a data frame

For simple filtering, `filter_match()` is usually enough. Let's keep only the `DEBUG` lines and convert our vector to a data frame.

```{r}
df <- 
  x |> 
  filter_match(str_detect(match, "^DEBUG")) |>
  as_tibble()

df
```

This gives one row per match and one column per field, which makes grouped summaries easy.

For example, we could create a custom summary per file for `DEBUG` lines where we see the number of matches and distinct matches as well as the first and last lines where we find a match.

```{r}
df |>
  summarise(
    n_matches = n(),
    n_distinct_matches = n_distinct(match),
    first_line = min(start_line),
    last_line = max(end_line),
    .by = path
  )
```

This kind of question is often easier to answer with a data frame than with the match vector directly.

## Filter with group-level information

A tabular workflow is also useful when filtering depends on group-level information.

For example, suppose we only want to keep the latest occurrence of each distinct `DEBUG` log message in each file.

```{r}
latest <-
  df |>
  mutate(
    match_rank_from_end = row_number(desc(start_line)),
    .by = c(path, match)
  ) |>
  filter(match_rank_from_end == 1L)

latest
```

The extra column `match_rank_from_end` is useful for now, but it is not part of a `seekr_match` vector. `as_match()` will ignore extra columns and validate the required match fields when converting back.

## Update replacements 

Finally, it is also easy to prepare some replacements. Here, we create different replacements depending on the content of each match.

```{r}
latest <-
  latest |>
  mutate(
    replacement = case_when(
      str_detect(match, "Disk") ~ str_to_upper(match),
      str_detect(match, "config") ~ str_to_lower(match),
      .default = str_to_title(match)
    )
  )

latest |>
  count(match, replacement)
```

## Convert back to a seekr_match vector

Before replacing files, convert the data frame back to a `seekr_match` vector. The result can be printed again with replacement previews now that some replacements have been set.

```{r}
to_replace <- as_match(latest)
to_replace
```

Finally, our `seekr_match` vector can be used to replace only the selected matches in the files by their corresponding `replacement`.

```{r, eval = FALSE}
replace_files(to_replace)
```

```{r cleaning, include = FALSE, eval = FALSE}
unlink(example_dir, recursive = TRUE)
unlink(backup_dir, recursive = TRUE)
```
