---
title: "Working with text"
description: >
  How to use match_text() and replace_text() when you want to search and
  replace text without letting seekr manage file input and output.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The usual `seekr` workflow starts from files.

You call `seek()`, `seekr()`, or `match_files()` to search files and create a `seekr_match` vector. You inspect, filter, or update that vector. Then, if needed, you pass it to `replace_files()` to apply the selected replacements back to disk.

That workflow is useful when files are the source of truth and when you want `seekr` to handle file reading and writing for you.

But sometimes the text you want to search does not come directly from files, or you need more control over how text is read, transformed, and written. For example, text may come from a database, an API, a package object, a clipboard, a web request, or a custom file-reading function. You may also want to decide exactly what happens after replacement: update a database row, send the result to an API, write a custom SQL script, preserve a specific encoding, or use your own output format.

For those cases, `seekr` provides two lower-level counterparts to `match_files()` and `replace_files()`:

- `match_text()` searches one in-memory string and returns a `seekr_match` vector.
- `replace_text()` applies a `seekr_match` vector back to an in-memory string and returns the modified text.

These functions let you use `seekr` as a search-and-replacement engine for strings. You still get structured match objects, context printing, replacement previews, filtering, and safety checks, but you remain responsible for where the text comes from and where the modified text goes.

## Example: updating SQL procedures from a table

Imagine that you have extracted a few SQL procedures from a database.

In this example, the procedures are already available in a tibble. In a real workflow, this table could have been created from a database query, an API call, or any other source.

```{r, echo = TRUE, eval = FALSE}
library(seekr)
library(dplyr)
library(purrr)
```

```{r, include = FALSE, eval = TRUE}
library(seekr)
library(dplyr)
library(purrr)
options(width = 150)
```

```{r, echo = FALSE}
df <- tibble(
  name = c(
    "refresh_customer_scores",
    "refresh_offer_flags",
    "archive_old_events",
    "refresh_campaign_targets",
    "cleanup_temp_tables"
  ),
  body = c(
    "
CREATE OR REPLACE FUNCTION refresh_customer_scores()
RETURNS void AS $$
BEGIN
  UPDATE customer_scores
  SET old_score = old_score + 1;
END;
$$ LANGUAGE plpgsql;
",
    "
CREATE OR REPLACE FUNCTION refresh_offer_flags()
RETURNS void AS $$
BEGIN
  UPDATE offer_flags
  SET old_flag = TRUE
  WHERE active = TRUE;
END;
$$ LANGUAGE plpgsql;
",
    "
CREATE OR REPLACE FUNCTION archive_old_events()
RETURNS void AS $$
BEGIN
  DELETE FROM events
  WHERE event_date < CURRENT_DATE - INTERVAL '365 days';
END;
$$ LANGUAGE plpgsql;
",
    "
CREATE OR REPLACE FUNCTION refresh_campaign_targets()
RETURNS void AS $$
BEGIN
  UPDATE campaign_targets
  SET old_segment = 'active'
  WHERE last_seen >= CURRENT_DATE - INTERVAL '30 days';
END;
$$ LANGUAGE plpgsql;
",
    "
CREATE OR REPLACE FUNCTION cleanup_temp_tables()
RETURNS void AS $$
BEGIN
  DELETE FROM temp_work_table;
END;
$$ LANGUAGE plpgsql;
"
  )
)
```

```{r}
df
```

Suppose we want to rename some old SQL identifiers before reviewing and re-running the procedures.

The text is not coming from files, so we use `match_text()` instead of `match_files()`.

`match_text()` searches one string at a time. It also needs a `path`, which is used as a source identifier in the resulting `seekr_match` vector. Here, we use the procedure name as that identifier.

```{r}
df <- 
  df |>
  mutate(
    x = map2(
      name, 
      body, 
      \(name, body) match_text(
        text = body,
        path = name,
        pattern = "old_([a-z_]+)",
        replacement = "new_\\1"
      )
    ),
    .before = name
  )
```

The result is a list-column of `seekr_match` vectors.

```{r}
df
```

Each row still contains the original text and the matches found in that text. This is useful because we can now use ordinary data-frame operations to decide which procedures we want to keep.

For example, we can keep only procedures where at least one match was found.

```{r}
df <- 
  df |>
  filter(!map_lgl(x, is_empty))

df
```

We can also inspect all matches together by combining the list-column into one `seekr_match` vector.

```{r}
x <- reduce(df$x, c, .init = new_seekr_match())
print(x, context = 2L)
```

At this point, `x` is a regular `seekr_match` vector and can be inspected like any other search result.

```{r}
summary(x)
```


## Filtering before replacement

One advantage of the in-memory workflow is that the match vectors remain linked to the text they came from.

Suppose we decide that only some procedures should be updated. For example, we may want to update the 3 refresh procedures, but leave other procedures unchanged for now.

```{r}
df <- 
  df |>
  filter(grepl("^refresh_", name))

df
```


## Applying replacements in memory

We can now apply the selected replacements with `replace_text()`.

Unlike `replace_files()`, this does not write anything to disk. It simply returns a modified string and the updated text is now just another column.

```{r}
df <- 
  df |>
  mutate(
    body_replaced = map2_chr(
      body, 
      x, 
      \(body, x) replace_text(text = body, x = x)
    ), 
    .before = body
  )

df
```

This is the point of the in-memory workflow: `seekr` helped us find, inspect, filter, and safely apply replacements, but it did not decide what to do with the result.


## Writing a custom SQL script

Because the updated procedures are ordinary strings, we can write them however we want.

For example, we can combine the updated procedures into a single SQL script. We can also add comments, separators, or any other metadata that makes sense for the target system.

```{r}
sql_script <- 
  df |>
  mutate(
    script_block = paste0(
      "-- Procedure: ", name, " -------------------------------\n",
      body_replaced
    )
  ) |>
  pull(script_block)

cat(sql_script, sep = "\n\n")
```

The script could then be written to disk, copied into a SQL client, sent to a deployment tool, or reviewed manually.

```{r, eval = FALSE}
writeLines(sql_script, "updated_procedures.sql")
```


## Safety checks still apply

`replace_text()` follows the same safety principle as `replace_files()`.

A `seekr_match` vector records a hash of the text that was searched. When you call `replace_text()`, `seekr` checks that the text supplied to `replace_text()` is the same text that was used to create the matches.

This prevents replacements from being applied to stale positions.

For example, if the text is changed after matching but before replacement, `replace_text()` will fail rather than applying replacements to a string that no longer corresponds to the recorded matches.

```{r, eval = FALSE}
x <- match_text(
  text = "Original text with old_name.",
  path = "example",
  pattern = "old_name",
  replacement = "new_name",
  encoding = "UTF-8"
)

replace_text(text = "Modified text with the match: old_name.", x)
```

The safe workflow is the same as with files: if the text changed, search again.


## Why not use stringr directly?

For simple in-memory replacements, using `stringr` directly is often the right choice.

If you already know exactly what should be replaced and you want to replace every occurrence in a string, a direct call to `stringr::str_replace_all()` is simpler than creating a `seekr_match` vector.

The reason to use `match_text()` and `replace_text()` is when replacement is not just a single immediate operation.

With direct replacement, the search and replacement happen at the same time. This is efficient, but it leaves little room for inspection. You do not get a structured object containing all matches, their locations, surrounding context, and planned replacements. You also cannot easily print the matches with context, preview the replacements, filter out false positives, update replacements match by match, and then apply only the selected changes.

To get that kind of workflow with `stringr` alone, you would need to first find all matches, store their positions, extract context, keep track of replacements, filter the result, and then carefully apply the selected replacements back to the original text. In other words, you would start rebuilding much of what a `seekr_match` vector already represents.

`match_text()` and `replace_text()` are useful when you want the flexibility of working with strings in memory, but still want the structured, inspectable, and safe replacement workflow that `seekr` provides.


## Summary

Use `match_files()` and `replace_files()` when files are the source of truth and you want `seekr` to handle file reading and writing.

Use `match_text()` and `replace_text()` when you want to bring your own text into the `seekr` workflow and decide yourself what to do with the modified result.

This is useful when text comes from a database, an API, an object in memory, or a custom reading process. It is also useful when you need full control over output: writing a custom script, updating a database, preserving a specific encoding, or sending the result somewhere else.

The core idea is the same in both workflows: search results become `seekr_match` vectors that can be inspected, filtered, updated, and then applied safely.
